---
title: "Data Visualization"
---

## Introduction

### Prerequisites

## First steps 

### The `mpg` data frame

### Creating a ggplot

### A graphing template

### Exercises

```{r}
#| label: load-packages
#| include: false 
library(ggplot2)
```

#### Run ggplot(data = mpg). What do you see?

empty gray background

#### How many rows are in mpg? How many columns?

There are `r nrow(mpg)` rows and `r ncol(mpg)` cols.

#### What does the drv variable describe? Read the help for ?mpg to find out.

the type of drive train, where f = front-wheel drive, r = rear wheel drive, 4 = 4wd

#### Make a scatterplot of hwy vs cyl.


```{r scatter, echo=FALSE}
library(ggplot2)
ggplot(data = mpg, aes(x = cyl, y = hwy)) +
  geom_point() +
  xlab("Number of Cylinders") +
  ylab("Highway Miles (per gallon)")
```

#### What happens if you make a scatterplot of class vs drv? Why is the plot not useful?

```{r scatter-class-drv, echo=FALSE}
ggplot(data = mpg, aes(x = drv, y = class)) +
  geom_point() +
  xlab("Drivetrain") +
  ylab("Class of vehicle")
```

<div class="solve">Not useful, both variables are categorical, no continuous relationship exists between variables.</div>

#### 3.3.1 Exercises

1. What’s gone wrong with this code? Why are the points not blue?

```{r bad-map, eval=FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = "blue"))
```

By assigning color inside aes, color is mapped to a variable, since only a factor of len 1 is present ("blue"), it is mapped to the first color in the default palette, red.

2. Which variables in mpg are categorical? Which variables are continuous? (Hint: type ?mpg to read the documentation for the dataset). How can you see this information when you run mpg?

*Categorical*: Manufacturer, model, year, cyl, trans, drv, fl, class;
*Continuous*: disp, cty, hwy

Described under colname, inferred by type, for ex. chr is a character, and thus a categorical label. Year, although an int, is in this case categorical. Cyl is also an int but is categorical; there is no 4.5 cylinder car (unless it is having a serious mechanical failure).
  
3. Map a continuous variable to color, size, and shape. How do these aesthetics behave differently for categorical vs. continuous variables?

```{r map-explore, echo=TRUE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = class, size=cty, shape = class))
```

4. What happens if you map the same variable to multiple aesthetics?

The variables legends are combined. 

5. What does the stroke aesthetic do? What shapes does it work with? (Hint: use ?geom_point)

```{r stroke, echo=TRUE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = class, size=cty, shape = class),
             stroke = 2)
```

Changes the thickness of line border on points. 

6. What happens if you map an aesthetic to something other than a variable name, like aes(colour = displ < 5)? Note, you’ll also need to specify x and y.


```{r conditional-map, echo=FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color =  displ < 5)) 
```

Colors points meeting that condition in one color, points not meeting that condition in another. FALSE is first alphabetically, so receives first color in palette (red) and TRUE receives blue.


#### 3.5.1 Exercises

1. What happens if you facet on a continuous variable?

A facet is created for every occurrence of a unique continuous variable. 

2. What do the empty cells in plot with facet_grid(drv ~ cyl) mean? How do they relate to this plot?

```{r no-facet-grid, echo=FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = drv, y = cyl))
```

Empty cells (shown below) indicate that no data is present for that particular pair of variables, that is, there are no 4-wheel drive 5 cylinder vehicles, nor 4 or 5 cylinder rear wheel drive vehicles. This is reflected in the above plot, where no point is present for those variable combinations combinations. 
```{r facet-grid, echo=FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = drv, y = cyl)) +
  facet_grid(drv ~ cyl)
```

3. What plots does the following code make? What does . do?
```{r mystery-plot, execute=FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(drv ~ .)

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(. ~ cyl)
```

Facet grid uses a row by column ordering. In the first plot, it will show facets by drive (f, r, 4) by row. The second plot will show cylinder count by column. The . is a dummy variable indicating no data. The code will not run without the dummy variable.

4. Take the first faceted plot in this section:

```{r facet-advantage, execute=FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ class, nrow = 2)
```

What are the advantages to using faceting instead of the colour aesthetic? What are the disadvantages? How might the balance change if you had a larger dataset?

Groups differentiated by color become difficult to discern when using more than 6 or more colors. Additionally, poor color choices can result in an inability to discern differences by group for those with colorblindness. Facets provide an unambiguous way of separating data by a categorical variable. It can make a visual comparison more difficult, which can be resolved by keeping each variable aligned along a common orientation (row vs column). With a larger number of variables, facets may become cumbersome. 

5. Read ?facet_wrap. What does nrow do? What does ncol do? What other options control the layout of the individual panels? Why doesn’t facet_grid() have nrow and ncol arguments?

6. When using facet_grid() you should usually put the variable with more unique levels in the columns. Why?
