---
title: "Explore - R for Data Science"
author: "Pete Lawson"
date: "2022-07-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
```

# Explore

Exercises for **Explore** section of the [R for Data Science](https://r4ds.had.co.nz/) textbook.

## Data Visualization

### First Steps 

#### 3.2.4 Exercises

1. Run ggplot(data = mpg). What do you see?

- empty gray background

2. How many rows are in mpg? How many columns?

- There are `r nrow(mpg)` rows and `r ncol(mpg)` cols.

3. What does the drv variable describe? Read the help for ?mpg to find out.

- the type of drive train, where f = front-wheel drive, r = rear wheel drive, 4 = 4wd

4. Make a scatterplot of hwy vs cyl.


```{r scatter, echo=FALSE}
ggplot(data = mpg, aes(x = cyl, y = hwy)) +
  geom_point() +
  xlab("Number of Cylinders") +
  ylab("Highway Miles (per gallon)")
```

5. What happens if you make a scatterplot of class vs drv? Why is the plot not useful?

```{r scatter-class-drv, echo=FALSE}
ggplot(data = mpg, aes(x = drv, y = class)) +
  geom_point() +
  xlab("Drivetrain") +
  ylab("Class of vehicle")
```

- Not useful, both variables are categorical, no continuous relationship exists between variables.

#### 3.3.1 Exercises

1. What’s gone wrong with this code? Why are the points not blue?

```{r bad-map, eval=FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = "blue"))
```

- By assigning color inside aes, color is mapped to a variable, since only a factor of len 1 is present ("blue"), it is mapped to the first color in the default palette, red.

2. Which variables in mpg are categorical? Which variables are continuous? (Hint: type ?mpg to read the documentation for the dataset). How can you see this information when you run mpg?

- Categorical: Manufacturer, model, year, cyl, trans, drv, fl, class;
- Continuous: disp, cty, hwy
- Described under colname, inferred by type, for ex. chr is a character, and thus a categorical label. Year, although an int, is in this case categorical. Cyl is also an int but is categorical; there is no 4.5 cylinder car (unless it is having a serious mechanical failure).
  
3. Map a continuous variable to color, size, and shape. How do these aesthetics behave differently for categorical vs. continuous variables?

```{r map-explore, echo=TRUE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = class, size=cty, shape = class))
```

4. What happens if you map the same variable to multiple aesthetics?

- The variables legends are combined. 

5. What does the stroke aesthetic do? What shapes does it work with? (Hint: use ?geom_point)

```{r stroke, echo=TRUE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = class, size=cty, shape = class),
             stroke = 2)
```

- Changes the thickness of line border on points. 

6. What happens if you map an aesthetic to something other than a variable name, like aes(colour = displ < 5)? Note, you’ll also need to specify x and y.


```{r conditional-map, echo=FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color =  displ < 5)) 
```

- Colors points meeting that condition in one color, points not meeting that condition in another. FALSE is first alphabetically, so receives first color in pallete (red) and TRUE receives blue.